Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-03 WSSSPE1_report/WSSSPE1_summary_report.tex
Blanton summary and Huang summary for Implementing Policy.

Two of the papers in this group focus on specific facets of software
design that are important factors in a project's sustainability but
are often addressed only late in the scientific software development
cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API
governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity
models for software security. The other two papers discuss
implementation strategies for science from the perspective of
facilitating many or all facets of sustainanability-oriented software
design for science: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
large projects that have software infrastructue development built-in
with cross-training domain scientist PIs in software engineering best
practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
specialized software engineering skills could be turned into shared
instrumentation with low barriers to access.

@@

% Finally, we’ll look at some papers that discuss implementing sustainability in scientific software.

% Blanton and Lenhardt from the Renaissance Computing Institute discuss these issues from a user perspective.

% The authors focus on a point that has been brought up many times in this context:
% There is a tension between writing code that is good enough just to “get it done,” i.e. to publish a paper about scientific results obtained using software, 
% and “getting it right,” that is, developing software that is comprehensible to future users and reviewers.
% Just because the elevator panel works like this doesn’t mean it’s sustainable for the long run.
% We don’t have a way to validate that the software used in a paper is actually done right.
% The best way to get software designed correctly is to make sure best practices are considered from the start.

% The authors highlight two models for sustainable software, at different extremes:
% One is what they call “co-funding”:
% In these projects, usually large, multi-year collaborations, 
% there is equal emphasis on both the science and the software development. 
% Both are planned into the project from inception. 
% In the life sciences, the iPlant Collaborative, Galaxy Project, and Qiime are good examples of these sorts of large, well-designed projects.

% At the other extreme, they discuss “software carpentry”: 
% in this model, it’s assumed that the scientists themselves will write and maintain their code.
% Groups like Software Carpentry and ROpenSci assume that 
% scientists won’t have access to dedicated software engineering, 
% so they try to give them tools to use best practices in their own software development.

% There might be a middle ground here: a way to get the engineering expertise that large co-funded projects have to individual scientist-developers.
% Hilmar Lapp of NESCent and I discuss one such possibility in our paper, Software Engineering as Instrumentation for the Long Tail of Scientific Software.

>>>>

Two of the papers in this group focus on specific facets of software
design that are important factors in a project's sustainability but
are often addressed only late in the scientific software development
cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API
governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity
models for software security. The other two papers discuss
implementation strategies for science from the perspective of
facilitating many or all facets of sustainanability-oriented software
design: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
large projects that have software infrastructue development built-in,
with cross-training domain scientist PIs in software engineering best
practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
specialized software engineering skills could be turned into shared
instrumentation with low barriers to access.

@@

API governance and cybersecurity measures appropriate for a project's
maturity are all but two facets of sustainability-oriented software
development. Others include user-centered interface design, test
engineering, dependency management, and deployment operations. Each of
these facets require specialized skills and training in software
engineering. How can the implementation of best practices and
guidelines along many or all of these different facets be facilitated
in scientific software development? Blanton and
Lenhardt~\cite{Blanton_WSSSPE} contrast two models. In one, the time
and personnel devoted to software engineering is ``co-funded'' with
the driving science project. This typically implies large multi-year
collaborative projects that to succeed require significant software
infrastructure to be built, and which thus have the funding to support
one or several software engineers. The sustainability of such projects
then depends on sustaining the funding. In the other model, domain
scientists also take the role of software developers, whether by
necessity such as funding limitations, or due to cross-disciplinary
professional interests. For this to result in sustainable software
products, the domain scientists need to be (or become) cross-trained
in software engineering standards and best practices.

In practice, there are example for both extremes of the spectrum. For
example, in the life sciences the iPlant Collaborative~\cite{iPlant},
the Galaxy Project~\cite{Galaxy}, and Qiime~\cite{Qiime} are large
multi-year projects with significant software infrastructure funding
and deliverables. Typically though these are the exception rather than
the rule, and particularly in the long tail of science the
scientist-developer predominates. For providing these with better
training in software engineering best practices, initiatives such as
Software Carpentry~\cite{SoftwareCarpentry} have been especially
valuable, and Software Carpentry also shown how such training could
scale.

However, there may also be a middle ground between the two
extremes. Huang and Lapp~\cite{Huang_WSSSPE} propose a Center of
Excellence model that leverages economies of scale to make software
engineering experts and their skills accessible to the long tail of
science. As Huang and Lapp discuss, this model could effectively turn
the utilization of software engineering expertise from a complex human
resource recruitment and management challenge, to buying time, when
and to the extent needed, on shared instrumentation. There is
precedence to using such a model to lower access barriers for long
tail science, in particular for new experimental technologies.  For
example, although the acquisition and operation of next-generation
high-throughput DNA sequencers requires substantial investments of
money, time, and expertise, the establishment of ``core facilities''
on many university campuses has made these technologies accessible to
a wide swath of scientists, with transformative results for science.

One of the important conclusions from this group of papers is that creating sustainable software requires paying attention not to one or two, but to several different facets of software engineering, each with corresponding best practices and standards of excellence. Even if a science project requires and has funding for a full software engineering FTE, what the project really needs could be fractions of different engineers with different specialty training. The vast majority of long tail science projects lacks the funding for even one full software engineer, let alone one who combines expertise in all the applicable facets of engineering. Some scientists in the long tail will have the professional interests to cross-train enough in software engineering to be successful with creating sustainable software, but it is likely unrealistic to expect that of all or even the majority. This is where a software engineering center of excellence can provide a critical service.
